---
title: "406 Final Project Code"
author: "Lydia Rogers"
date: "3/24/2021"
output:
  pdf_document: default
  html_document: default
---

## Overview of Computations

1. Load data
2. Remove outcomes that are not of interest to us
3. Generate new indicator variables for black cats and pitbulls
4. Parameter estimation for adoption outcomes of black cats, pitbulls, all cats, and all dogs entering shelter
5. Least squares regression to predict length of stay of adopted animals from age, intake type, animal type, sex of animal

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(coin)
library(knitr)
```

```{r}
intake <- read_csv("Austin_Animal_Center_Intakes.csv")
outcome <- read_csv("Austin_Animal_Center_Outcomes.csv")
intake$DateTime <- mdy_hms(intake$DateTime)
intake$MonthYear <- mdy_hms(intake$MonthYear)
outcome$DateTime <- mdy_hms(outcome$DateTime)
outcome$MonthYear <- mdy_hms(outcome$MonthYear)
```

Combine datasets to create comprehensive overview
Create new variables: length of time spent at shelter, datetime age in months, black cat indicator, pitbull indicator. Also, remove outcomes that are not of interest to us.

Some differences in intakes and outcomes have negative values, indicating that a particular animal was at the shelter more than once (some animals are adopted and returned). For these animals, we keep only non-negative differences in outcome
```{r}
intake_outcome <- intake %>% left_join(outcome, by = c("Animal ID" = "Animal ID"))

intake_outcome <- intake_outcome %>% mutate(
  TimeAtShelter = difftime(DateTime.y, DateTime.x, units = c("days")), 
  BlackCat = `Animal Type.x` == "Cat" & Color.x == "Black", 
  Pitbull = str_detect(Breed.x, "Pit Bull"), 
  adopted = ifelse(`Outcome Type` == "Adoption", 1, 0)) %>% 
  filter(
    `Outcome Type` %in% c("Adoption", "Died", "Euthanasia", "Relocate", "Transfer", NA), 
    TimeAtShelter >= 0, 
    `Animal Type.x` %in% c("Dog", "Cat"))
```

Basic visualizations for exploratory data analysis for length of stay and outcomes for dogs and cats
```{r}
ggplot(data = filter(intake_outcome, TimeAtShelter < 100)) + 
  geom_density(aes(x = TimeAtShelter, color = `Animal Type.x`)) + 
  labs(title = "Cats and Dogs Spending Less than 100 Days at AAC, 2013-2021") 
#excludes 10,098 animals who spent more than 100 days at the shelter

ggplot(data = intake_outcome) + 
  geom_bar(aes(x = `Animal Type.x`, fill = `Outcome Type`)) + 
  labs(x = "Animal Type", title = "Outcomes of Interest for Cats and Dogs at AAC, 2013-2021")
```

## Questions 1 and 2: Basic analysis and permutation tests for differences between black cats and all other cats and pit bulls and all other dogs

* For all cats/dogs adopted from AAC in this time frame, how do the lengths of stays compare between black cats/pit bulls and all other cats/dogs?
* How do the proportions of black cats/pit bulls adopted vs. other outcomes compare with the same proportion amongst all other cats/dogs?

```{r}
#proportions of adoptions for black cats
prop.table(table(intake_outcome$BlackCat, intake_outcome$`Outcome Type`), 1)

#proportions of adoptions for pit bulls
prop.table(table(intake_outcome$Pitbull, intake_outcome$`Outcome Type`), 1)

#plot length of time at shelter for adopted cats
ggplot(data = filter(intake_outcome, `Outcome Type` == "Adoption", `Animal Type.x` == "Cat", TimeAtShelter < 100)) + 
  geom_density(aes(x = TimeAtShelter, color = BlackCat))

#plot length of time at shelter for adopted dogs
ggplot(data = filter(intake_outcome, `Outcome Type` == "Adoption", `Animal Type.x` == "Dog", TimeAtShelter < 100)) + geom_density(aes(x = TimeAtShelter, color = Pitbull))

#fisher exact test for difference in proportions between all cats and balck cats
cats <- filter(intake_outcome, `Animal Type.x` == "Cat")
fisher.test(cats$adopted, cats$BlackCat)

#fisher exact test for difference in proportions between all dogs and pit bulls
dogs <- filter(intake_outcome, `Animal Type.x` == "Dog")
fisher.test(dogs$adopted, dogs$Pitbull)
```

Permutation test for difference in mean length of time spent at AAC for black cats and all other cats who were eventually adopted
```{r}
adopted_cats <- intake_outcome %>% filter(`Outcome Type` == "Adoption", `Animal Type.x` == "Cat")

actual_diff <- diff(by(adopted_cats$TimeAtShelter, adopted_cats$BlackCat, median))

dist <- replicate(10000, diff(by(adopted_cats$TimeAtShelter, sample(adopted_cats$BlackCat, length(adopted_cats$BlackCat), FALSE), median)))

hist(dist, col = "black", xlim = c(-2, 8), breaks = 100, main = "Permutated and Actual Differences in Medians 
     Between Black Cats and Other Cats", xlab = "Difference in Medians")
abline(v = actual_diff, col = "blue", lwd = 2)
sum(dist > actual_diff)/10000
```

Next, do the same thing for pit bulls
```{r}
adopted_dogs <- intake_outcome %>% filter(`Outcome Type` == "Adoption", `Animal Type.x` == "Dog")

actual_diff <- diff(by(adopted_dogs$TimeAtShelter, adopted_dogs$Pitbull, median))

dist <- replicate(10000, diff(by(adopted_dogs$TimeAtShelter, sample(adopted_dogs$Pitbull, length(adopted_dogs$Pitbull), FALSE), median)))

hist(dist, xlim = c(-2, 29), breaks = 100, main = "Permutated and Actual Differences in Medians 
     Between Pit Bulls and Other Dogs", xlab = "Difference in Medians")
abline(v = actual_diff, col = "blue", lwd = 2)
sum(dist > actual_diff)/10000

```

Finally, perform least squares regression to predict length of stay for cats and dogs
```{r}
adopted <- mutate(adopted, young = !str_detect(`Age upon Outcome`, "year"))

mod <- lm(as.numeric(TimeAtShelter) ~ BlackCat + Pitbull + `Intake Type` + young + `Sex upon Outcome` + `Animal Type.y`,data = adopted)

summary(mod)

knitr::kable(data.frame(mod$coefficients, summary(mod)$coefficients[,4]), col.names = c("Coefficient Estimate", "P-value"))
```

## Simulated Data
```{r}
probs <- c(0.01, 0.1, 0.25, 0.5, 0.75)

# new samples for simulation
sim_cats1 <- rbinom(nrow(cats), 1, probs[1])
sim_cats2 <- rbinom(nrow(cats), 1, probs[2])
sim_cats3 <- rbinom(nrow(cats), 1, probs[3])
sim_cats4 <- rbinom(nrow(cats), 1, probs[4])
sim_cats5 <- rbinom(nrow(cats), 1, probs[5])

sim_dogs1 <- rbinom(nrow(dogs), 1, probs[1])
sim_dogs2 <- rbinom(nrow(dogs), 1, probs[2])
sim_dogs3 <- rbinom(nrow(dogs), 1, probs[3])
sim_dogs4 <- rbinom(nrow(dogs), 1, probs[4])
sim_dogs5 <- rbinom(nrow(dogs), 1, probs[5])

#fisher's exact test on new samples
fisher.test(cats$adopted, sim_cats1)$p.value
fisher.test(cats$adopted, sim_cats2)$p.value
fisher.test(cats$adopted, sim_cats3)$p.value
fisher.test(cats$adopted, sim_cats4)$p.value
fisher.test(cats$adopted, sim_cats5)$p.value

fisher.test(dogs$adopted, sim_dogs1)$p.value
fisher.test(dogs$adopted, sim_dogs2)$p.value
fisher.test(dogs$adopted, sim_dogs3)$p.value
fisher.test(dogs$adopted, sim_dogs4)$p.value
fisher.test(dogs$adopted, sim_dogs5)$p.value
```
analyze usefulness of linear model
```{r}
sum((mod$residuals)^2)
```


### Some code used to intial EDA for the initial draft of the paper
```{r}
#outcome <- outcome %>% mutate(black_cat = (Color == "Black" & `Animal Type` == "Cat"), )
#black.outcome <- table(outcome$`Outcome Type`, outcome$black)
#sum(outcome$black)
#nrow(outcome)
#prop.table(black.outcome, 2)

#black.cat.outcome <- table(outcome$`Outcome Type`, outcome$black_cat)
#prop.table(black.cat.outcome, 2)
```

```{r}
#black cats vs all cats
#cats_outcome <- outcome %>% filter(`Animal Type` == "Cat")
#prop.table(table(cats_outcome$`Outcome Type`, cats_outcome$black), 2)
```

```{r}
#find all cats from intake and see which were adopted
#intake <- intake %>% mutate(black = Color == "Black")
#combined_intake_outcome <- left_join(intake, outcome, by=c("Animal ID", "Breed", "Color"))
#prop.table(table(combined_intake_outcome$`Outcome Type`, combined_intake_outcome$black), 2)
#sum(is.na(combined_intake_outcome$`Outcome Type`))


```

```{r}
# plot data of cats vs. dogs
#ggplot(data = filter(combined_intake_outcome, `Animal Type.x` == "Cat" | `Animal Type.x` == "Dog")) + geom_bar(aes(x = `Animal Type.x`, fill = `Outcome Type`))
```